Haul-out behavior and aerial survey detectability of seals in the Bering and Chukchi seas

### 2022-03-07 analysis/paper/London_HauloutBeahvior.Rmd
add section for Supplemental Material
apply edits from Jim Lee (AFSC review)

    address: Marine Mammal Laboratory, Alaska Fisheries Science Center, NOAA Fisheries, Seattle, Washington, USA

@@

The remoteness of the Bering and Chukchi seas means direct scientific observation
of seal behavior is impractical. Thus, bio-logging devices are especially useful
tools for collecting key information on movement and haul-out behavior for these
species. Bio-logging records of time spent out of the water provide valuable
data for identifying covariates that explain variation in haul-out behavior. For
instance, in the Antarctic, Bengtson and Cameron [-@bengtson2004] relied on
bio-logging data to demonstrate greater haul-out propensity in juvenile
crabeater seals (_Lobodon carcinophaga_) than adults, with highest probabilities
in February and at times close to solar noon. In the Arctic, Von Duyke et al.
[-@vonduyke2020] used satellite-linked bio-loggers to corroborate seasonal
changes between diurnal and nocturnal haul-out behavior of ringed seals
previously described by Kelly and Quakenbush [-@kelly1990c] using VHF radio tags
and direct observation. Bengtson et al. [-@bengtson2005a] documented a higher
propensity for ringed seal basking near solar noon, as did Ver Hoef et al.
[-@verhoef2014a] in an analysis of bearded, ribbon, and spotted seals using much
larger sample sizes. Olnes et al. [-@olnes2020] showed that the proportion of
time bearded seals spent hauled out progressively increased through spring and
summer. And, Ver Hoef et al. [-@verhoef2014a] found haul-out probabilities
increased gradually starting in March and peaking in May and June for bearded,
ribbon, and spotted seals.

@@

Data used in our study came from bio-loggers deployed on bearded, ribbon, and
spotted seals in the Bering, Chukchi, and western Beaufort seas by multiple
organizations as part of collaborative investigations from 2005 through 2020.
Seals were captured using nets and bio-loggers were attached during studies based
in coastal Alaska communities or on research ships. Ship-based capture events
occurred during spring near the southern ice edge in the Bering Sea. Land-based
capture events occurred from May to October, generally between the coastal
communities of Scammon Bay, Alaska in the Bering Sea, Utqiaġvik, Alaska in the
Chukchi Sea, and Nuiqsut in the Beaufort Sea from 2005 to 2020. We refer readers
to the primary literature for detailed capture and bio-logger attachment methods
(Supplemental Material, Table).

@@

Of key interest in this study was the relationship between haul-out behavior and
weather covariates that vary with time and seal location. We explored the use
of a continuous-time correlated random walk [@johnson2008a] movement model
to predict locations at specific times. However, the sparse nature of data from
some bio-loggers, especially those mounted to the rear flipper, resulted in poor
modeling performance or convergence issues. For this study, we calculated a
weighted average daily location where the inverse of the estimated Argos or
FastLoc GPS location error was used for the weight. Each Argos location estimate
was assigned an error radius based on either the categorical location quality (
_3_=250m, _2_=500m, _1_=1500m, _0_=2500m [@lopez2013]; we chose 2500m for
location classes _A_ and _B_) or, when available, the estimated error radius
from the Argos Kalman filter algorithm. Location estimates from FastLoc GPS were
all assigned an error radius of 50m. On days when haul-out observations were
present but location data were missing we used the seal's last calculated weighted
average daily location; days when the location intersected with land
were removed from the seal's record. We recognized that bearded and spotted
seals haul out on land. However, assessing the relationship between haul-out
behavior and weather covariates and seals' availability for aerial surveys on
land was outside the scope of this study. Additionally, any daily locations on
land are likely more reflective of coordinate averaging and measurement error,
rather than actual use of coastal features. Figure \@ref(fig:dataMap) shows the
spatial distribution of weighted locations with available haul-out behavior data
used for analysis of each species across the study area.

@@

In addition to sex and age class, we analyzed variables that might help explain
variation in haul-out probabilities. These included day-of-year (for seasonal
effects) and local solar hour (for diurnal effects). Solar hour was calculated
using the {solaR} package [@perpinan2012] within the R statistical environment
[@rcoreteam2021] using the daily locations. We also linked the weighted average
daily locations from our bio-loggers to key weather values from the North
American Regional Reanalysis (NARR) model produced by the National Centers for
Environmental Prediction [@mesinger2006]. The NARR model assimilates
observational data to produce a long-term picture of weather over North America
and portions of the surrounding seas. Weather variables are made
available across the region 8 times daily. For this study, NARR weather values
were subset to the extent of our study area over the Bering and Chukchi seas at
3-hr intervals based on the native grid resolution of 32km (1024km^2^). The
following meteorological variables are known to affect haul-out behavior in
other Arctic pinnipeds [@reder2003;
@udevitz2009; @perry2017] and were interpolated and assigned to daily seal
locations using a bilinear method: 1) air temperature at 2m above the earth's
surface, 2) wind consisting of northerly and easterly vector components
converted to wind speed using the Euclidean norm, 3) barometric pressure at sea
level, and 4) precipitation (Table \@ref(tab:tblCovariates)).

@@

Notably missing from this list of explanatory variables is any spatial-temporal
representation of sea ice concentration, area, or extent. This may seem
counterintuitive when modeling the haul-out behavior of seal species with such a
close association to sea ice; seals haul out in the presence of sea ice, and we
could assess the local concentration of sea ice during these events (see
[@olnes2020]). However, a major focus of this study is to develop models
applicable for aerial survey correction factors and using sea ice as a covariate
would almost certainly bias haul-out predictions towards those seals that are on
or near ice and therefore more likely to haul out. Since aerial surveys can only
detect seals on ice, abundance estimates would be missing a correction for those
seals that are away from ice (e.g., on foraging trips). Additionally, many of
the deployments consisted of a single device attached to the rear flipper of the
seal which do not provide at-sea locations, limiting our ability to fully
evaluate fine-scale habitat preferences related to sea-ice. Lastly, our study
was limited to the spring season when seal haul-out tendencies are strongly
influenced by pupping, nursing, breeding behavior, and molt and these drivers
are likely more influential than specific sea-ice concentration. Crawford et al
[-@crawford2019] compared haul-out probability models for ringed seals and found
those that only included season (and not sea-ice concentration) were the most
parsimonious. For these reasons, we have elected not to use sea ice
concentration as a predictor for haul-out probability in the present study.

@@

We also investigated the influence of weather on haul-out probabilities,
including wind speed, temperature, barometric pressure, precipitation, and wind
chill. These have been investigated for walruses (e.g. Udevitz et al.
[-@udevitz2009]) and a few select studies of ice-associated seals
[@hamilton2018, @perry2017]. In our study, ribbon seals seemed to be the most
influenced by weather, with wind, temperature, and barometric pressure all being
important components of the model. Spotted seals were most affected by wind and
barometric pressure. And, for bearded seals, the model indicated wind and
temperature had the greatest impact. In general, and as might be expected, seals
were more likely to haul-out when daily temperatures were warmer, winds speeds
were lower, barometric pressure was higher, and precipitation was lower. Those
weather conditions are general indicators of increased solar radiation and lower
convective heat loss, both of which provides energetic benefits. Low winds and
precipitation could also enhance predator detection. Our results highlight the
importance of incorporating weather covariates when analyzing haul-out behavior
and calculating availability corrections for aerial surveys.

@@

# Acknowledgements {-}

@@

We are grateful for the assistance in catching and sampling seals by Ryan Adam,
James Bailey, Michelle Barbieri, John Bengtson, Gavin Brady, Vladamir Burkanov,
Cynthia Christman, Sarah Coburn, Shawn Dahle, Rob Delong, Stacy DiRocco, Deb
Fauquier, Shannon Fitzgerald, Kathy Frost, Scott Gende, Tracey Goldstein, Jeff
Harris, Jason Herreman, Markus Horning, John Jansen, Shawn Johnson, Charles
Littnan, Lloyd Lowry, Brett McClintock, Erin Moreland, Mark Nelson, Justin
Olnes, Lorrie Rea, Bob Shears, Gay Sheffield, Brent Stewart, Dave Withrow, and
Heather Ziel. We also appreciate the commitment to science and safety by all
officers and crew of the NOAA Ship Oscar Dyson, the NOAA Ship MacArthur II, and
the RV Thomas G. Thompson.

>>>>

    address: Marine Mammal Laboratory, Alaska Fisheries Science Center, National Marine Fisheries Service, NOAA, Seattle, Washington, USA

@@

The remoteness of the Bering and Chukchi seas means direct scientific observation
of seal behavior is impractical. Thus, bio-logging devices are especially useful
tools for collecting key information on movement and haul-out behavior for these
species. Bio-logging records of time spent out of the water provide valuable
data for identifying covariates that explain variation in haul-out behavior. For
instance, in the Antarctic, Bengtson and Cameron [-@bengtson2004] relied on
bio-logging data to demonstrate greater haul-out propensity in juvenile
crabeater seals (_Lobodon carcinophaga_) than adults, with highest probabilities
in February and at times close to solar noon. In the Arctic, Von Duyke et al.
[-@vonduyke2020] used satellite-linked bio-loggers to corroborate seasonal
changes between diurnal and nocturnal haul-out behavior of ringed seals
previously described by Kelly and Quakenbush [-@kelly1990c] using VHF radio tags
and direct observation. Bengtson et al. [-@bengtson2005a] documented a higher
propensity for ringed seal basking near solar noon, as did Ver Hoef et al.
[-@verhoef2014a] in an analysis of bearded, ribbon, and spotted seals using much
larger sample sizes. Olnes et al. [-@olnes2020] showed that the proportion of
time bearded seals spent hauled out progressively increased through spring and
summer, and Ver Hoef et al. [-@verhoef2014a] found haul-out probabilities
increased gradually starting in March and peaked in May and June for bearded,
ribbon, and spotted seals.

@@

Data used in our study came from bio-loggers deployed on bearded, ribbon, and
spotted seals in the Bering, Chukchi, and western Beaufort seas by multiple
organizations as part of collaborative investigations from 2005 through 2020.
Seals were captured using nets and bio-loggers were attached during studies based
in coastal Alaska communities or on research ships. Ship-based capture events
occurred during spring near the southern ice edge in the Bering Sea. Land-based
capture events occurred from May to October, generally between the coastal
communities of Scammon Bay, Alaska in the Bering Sea, Utqiaġvik, Alaska in the
Chukchi Sea, and Nuiqsut in the Beaufort Sea from 2005 to 2020. We refer readers
to the primary literature for detailed capture and bio-logger attachment methods
(see Supplemental Material, Table 1).

@@

Of key interest in this study was the relationship between haul-out behavior and
weather covariates that vary with time and seal location. We explored the use
of a continuous-time correlated random walk [@johnson2008a] movement model
to predict locations at specific times. However, the sparse nature of data from
some bio-loggers, especially those mounted to the rear flipper, resulted in poor
modeling performance or convergence issues. For this study, we calculated a
weighted average daily location where the inverse of the estimated Argos or
FastLoc GPS location error was used for the weight. Each Argos location estimate
was assigned an error radius based on either the categorical location quality (
_3_ = 250 m, _2_ = 500 m, _1_ = 1500 m, _0_ = 2500 m [@lopez2013]; we chose 2500 m for
location classes _A_ and _B_) or, when available, the estimated error radius
from the Argos Kalman filter algorithm. Location estimates from FastLoc GPS were
all assigned an error radius of 50 m. On days when haul-out observations were
present but location data were missing we used the seal's last calculated weighted
average daily location; days when the location intersected with land
were removed from the seal's record. We recognized that bearded and spotted
seals haul out on land. However, assessing the relationship between haul-out
behavior and weather covariates and seals' availability for aerial surveys on
land was outside the scope of this study. Additionally, any daily locations on
land are likely more reflective of coordinate averaging and measurement error,
rather than actual use of coastal features. Figure \@ref(fig:dataMap) shows the
spatial distribution of weighted locations with available haul-out behavior data
used for analysis of each species across the study area.

@@

In addition to sex and age class, we analyzed variables that might help explain
variation in haul-out probabilities. These included day-of-year (for seasonal
effects) and local solar hour (for diurnal effects). Solar hour was calculated
using the {solaR} package [@perpinan2012] within the R statistical environment
[@rcoreteam2021] using the daily locations. We also linked the weighted average
daily locations from our bio-loggers to key weather values from the North
American Regional Reanalysis (NARR) model produced by the National Centers for
Environmental Prediction [@mesinger2006]. The NARR model assimilates
observational data to produce a long-term picture of weather over North America
and portions of the surrounding seas. Weather variables are made
available across the region 8 times daily. For this study, NARR weather values
were subset to the extent of our study area over the Bering and Chukchi seas at
3-hr intervals based on the native grid resolution of 32 km (1024 km^2^). The
following meteorological variables are known to affect haul-out behavior in
other Arctic pinnipeds [@reder2003;
@udevitz2009; @perry2017] and were interpolated and assigned to daily seal
locations using a bilinear method: 1) air temperature at 2 m above the Earth's
surface, 2) wind consisting of northerly and easterly vector components
converted to wind speed using the Euclidean norm, 3) barometric pressure at sea
level, and 4) precipitation (Table \@ref(tab:tblCovariates)).

@@

Notably missing from this list of explanatory variables is any spatial-temporal
representation of sea ice concentration, area, or extent. This may seem
counterintuitive when modeling the haul-out behavior of seal species with such a
close association to sea ice; seals haul out in the presence of sea ice, and we
could assess the local concentration of sea ice during these events (see
[@olnes2020]). However, a major focus of this study is to develop models
applicable for aerial survey correction factors and using sea ice as a covariate
would almost certainly bias haul-out predictions towards those seals that are on
or near ice and therefore more likely to haul out. Since aerial surveys can only
detect seals on ice, abundance estimates would be missing a correction for those
seals that are away from ice (e.g., on foraging trips). Additionally, many of
the deployments consisted of a single device attached to the rear flipper of the
seal which do not provide at-sea locations, limiting our ability to fully
evaluate fine-scale habitat preferences related to sea-ice. Lastly, our study
was limited to the spring season when seal haul-out tendencies are strongly
influenced by pupping, nursing, breeding behavior, and molt and these drivers
are likely more influential than specific sea-ice concentration. Crawford et al.
[-@crawford2019] compared haul-out probability models for ringed seals and found
those that only included season (and not sea-ice concentration) were the most
parsimonious. For these reasons, we have elected not to use sea ice
concentration as a predictor for haul-out probability in the present study.

@@

We also investigated the influence of weather on haul-out probabilities,
including wind speed, temperature, barometric pressure, precipitation, and wind
chill. These have been investigated for walruses (e.g. Udevitz et al.
[-@udevitz2009]) and a few select studies of ice-associated seals
[@hamilton2018, @perry2017]. In our study, ribbon seals seemed to be the most
influenced by weather, with wind, temperature, and barometric pressure all being
important components of the model. Spotted seals were most affected by wind and
barometric pressure. For bearded seals, the model indicated wind and
temperature had the greatest impact. In general, and as might be expected, seals
were more likely to haul-out when daily temperatures were warmer, winds speeds
were lower, barometric pressure was higher, and precipitation was lower. Those
weather conditions are general indicators of increased solar radiation and lower
convective heat loss, both of which provides energetic benefits. Low winds and
precipitation could also enhance predator detection. Our results highlight the
importance of incorporating weather covariates when analyzing haul-out behavior
and calculating availability corrections for aerial surveys.

@@

# Acknowledgments {-}

@@

We are grateful for the assistance in catching and sampling seals by Ryan Adam,
James Bailey, Michelle Barbieri, John Bengtson, Gavin Brady, Vladamir Burkanov,
Cynthia Christman, Sarah Coburn, Shawn Dahle, Rob Delong, Stacy DiRocco, Deb
Fauquier, Shannon Fitzgerald, Kathy Frost, Scott Gende, Tracey Goldstein, Jeff
Harris, Jason Herreman, Markus Horning, John Jansen, Shawn Johnson, Charles
Littnan, Lloyd Lowry, Brett McClintock, Erin Moreland, Mark Nelson, Justin
Olnes, Lorrie Rea, Bob Shears, Gay Sheffield, Brent Stewart, Dave Withrow, and
Heather Ziel. We also appreciate the commitment to science and safety by all
officers and crew of the NOAA ship _Oscar Dyson_, the NOAA ship _MacArthur II_, and
the RV _Thomas G. Thompson_.
